Collective excitations in competing phases in two and three dimensions

### 2024-02-19 paper/main.tex
plot formatting
Feierabend

\author{Joshua Alth\"user}\email{joshua.althueser@tu-dortmund.de}
\affiliation{Condensed Matter Theory, TU Dortmund University,
Otto-Hahn Stra\ss{}e 4, 44221 Dortmund, Germany}

\author{G\"otz S.~Uhrig}
\email{goetz.uhrig@tu-dortmund.de}
\affiliation{Condensed Matter Theory, TU Dortmund University,
Otto-Hahn Stra\ss{}e 4, 44221 Dortmund, Germany}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                  Abstract                                                 %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
    We investigate the superconducting (SC), charge-density wave (CDW), and antiferromagnetic (AFM) phases 
    in the extended Hubbard model at zero temperature and half-filling.
    We employ the iterated equations of motion approach \cite{Kalthoff17,bleicker18} to compute the two-particle Green's functions and by extension, the corresponding spectral densities.
    This renders comprehensive analysis of the behavior of collective excitations possible as the model's parameters are changed across phase transitions.
    We identify the well-known amplitude (Higgs) and phase (Anderson-Bogoliubov) modes within the superconducting phase and observe a similar excitation in the CDW phase which shifts towards zero energy as the system approaches the phase transition to the SC phase.
    In the CDW phase, close to the phase transition to the AFM phase, we find a collective mode that does not change significantly and another mode that becomes soft as the phase boundary is approached.
\end{abstract}

@@

The study of collective excitations is of great scientific interest as it sheds light on the intricate dynamics of correlated electron systems, providing crucial insights into emergent material properties.
On the theoretical side, the Hubbard model has been employed in a plethora of previous studies. 
Early studies proved the existence of eigenstates to the Hubbard Hamiltonian that exhibit off-diagonal long-range order,
encouraging the model's usage for the description of high-temperature superconductivity \cite{yang89}.
Shortly afterward, an exact SO(4) symmetry was discovered, which allows a degeneracy of superconductivity (SC) and a charge-density wave (CDW) governed by an attractive on-site interaction \cite{yang90}.
This coexistence can be argued by the existence of a particle-hole transformation on bipartite lattices, 
that maps the attractive Hubbard model exactly onto the repulsive one, exhibiting antiferromagnetism (AFM) \cite{Hirsch85}.
The previously mentioned SC and CDW phases map to different spin operators \cite{zitko15,lieb89}.

Numerous studies have investigated the phases and various quantities of the Hubbard model in equilibrium systems, including various additional interactions with and without doping.
\cite{Micnas88,Micnas88b,Micnas89,Dzierzawa92,Kostyrko92,Eriksson95,Staudt00,Onari04,Toschi05,Brackett16,Paki19,romer20,Sushchyev22}.
For a long time, research on the dynamics of the superconducting gap parameter focused on its behavior after an instantaneous quench, 
yielding damped or persistent oscillations depending on the magnitude of the quench \cite{Volkov73,Yuzbashyan05,Yuzbashyan06,Barankov06,Cui19}.
More recent experimental and theoretical studies examined the driven systems in the hope of inducing superconductivity \cite{Nicoletti14,Krull14,Moor14,Casandruc15,patel16,sentef17,Buenemann17}.

In this paper, we will restrict ourselves to the half-filled Hubbard model, including an additional intersite interaction, on a square and a simple cubic lattice.
The 2D case already exhibits a wide range of possible phases, including CDW, AFM, $s$- and $d_{x^2 - y^2}$-wave superconductivity, and phase-separated states \cite{Micnas88b,Tsuchiura95,Su01,Su04,ha11,Huang13,Jiang22,Linner23}.
We expand these well-known phase diagrams to the simple cubic lattice where we find a similar structure for CDW, AFM, and $s$-wave SC.

@@

Moreover, we demonstrate an explicit way to obtain various Green's functions from the aforementioned methodology.
By extension, we also obtain the spectral functions of the investigated systems and discuss the excitations found therein.
The prominent examples in the SC phase are the well-known phase (Anderson-Bogoliubov) and amplitude (Higgs) modes \cite{Kulik1981,Varma02,Cea14,Measson14,Tsuji15,Krull16,Schwarz20,Fan22}.
The former is located at zero energy as our model describes neutral superfluids.
The latter is found at the lower edge of the two-particle continuum.

@@

In our study, we employ the extended Hubbard model as it encompasses assorted phases and thereby provides direct access to the rich excitation spectra therein.
Its Hamiltonian is given by
\begin{equation}
    \label{eqn:full_hamiltonian}
    \begin{aligned}
        H = &-t \sum_{\langle i, j \rangle, \sigma} \left( c_{i\sigma}^\dagger c_{j\sigma} + \text{h.c.} \right) 
        + \mu \sum_{i,\sigma} n_{i\sigma} \\
        &+ U \sum_{i} n_{i\uparrow} n_{i\downarrow} 
        + \frac{V}{2} \sum_{\langle i, j\rangle, \sigma} n_{i\sigma} n_{j\sigma}\,,
    \end{aligned}
\end{equation}
where $c_{i\sigma}^{(\dagger)}$ annihilates (creates) an electron with spin $\sigma$ on lattice site $i$ 
and $\langle i, j\rangle$ denotes the summation over the next neighbors.
The parameters are the hopping amplitude $t$, the onsite interaction $U$, the intersite interaction $V$, and the chemical potential $\mu$.
Applying a Fourier transform into $k$ space yields the single-particle dispersion 
\begin{equation}
    \epsilon_0 (\vk) = -2t \sum_{\alpha=1}^D \cos(k_\alpha)\,,\quad k_\alpha \in [-\pi, \pi)\,,
\end{equation}
with the system's dimension $D$ and the dimensionless momentum $\vk$.

@@

In this article, we will be using the operator pseudo scalar product
\begin{equation}
\label{eqn:scalar_product}
    (A | B) \coloneqq  \langle [A^\dagger, B] \rangle\,,
\end{equation}
where the expectation values are taken with respect to the thermal equilibrium of the mean-field Hamiltonian \eqref{eqn:mf_hamiltonian}.
Note, that this is not a proper scalar product, as a norm defined using it can also be negative.
For example, assume that for some operator $A$
\begin{equation}
    ||A|| = (A | A) = \langle [A^\dagger, A] \rangle > 0\,,
\end{equation}
then
\begin{equation}
    ||A^\dagger|| = (A^\dagger | A^\dagger) = \langle [A, A^\dagger] \rangle = - ||A|| < 0\,.
\end{equation}
Nevertheless, this does not pose any issue in our calculations.
We will compute $\mM$ by commutating with the full Hubbard Hamiltonian \eqref{eqn:full_hamiltonian} enabling us to capture collective behavior rather than the one particle dynamics given by the mean-field approximation.

>>>>

\author{Joshua Alth\"user}\email{joshua.althueser@tu-dortmund.de}
\affiliation{Condensed Matter Theory, TU Dortmund University,
Otto-Hahn Stra\ss{}e 4, 44227 Dortmund, Germany}

\author{G\"otz S.~Uhrig}
\email{goetz.uhrig@tu-dortmund.de}
\affiliation{Condensed Matter Theory, TU Dortmund University,
Otto-Hahn Stra\ss{}e 4, 44227 Dortmund, Germany}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                  Abstract                                                 %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
    We investigate the superconducting (SC), charge-density wave (CDW), and antiferromagnetic (AFM) phases 
    in the extended Hubbard model at zero temperature and half-filling.
    We employ the iterated equations of motion approach \cite{Kalthoff17,bleicker18} to compute the two-particle Green's functions and their spectral densities.
    This renders a comprehensive analysis of the behavior of collective excitations possible as the model's parameters are changed across phase transitions.
    We identify the well-known amplitude (Higgs) and phase (Anderson-Bogoliubov) modes within the superconducting phase and observe a similar excitation in the CDW phase which shifts towards zero energy as the system approaches the phase transition to the SC phase.
    In the CDW phase, close to the phase transition to the AFM phase, we find a collective mode that does not change significantly and another mode that becomes soft as the phase boundary is approached.
\end{abstract}

@@

The study of collective excitations is of great interest as it sheds light on the intricate dynamics of correlated electron systems, providing crucial insight into emergent material properties.
On the theoretical side, the Hubbard model has been employed in a plethora of previous studies. 
Early studies proved the existence of eigenstates to the Hubbard Hamiltonian that exhibit off-diagonal long-range order,
encouraging the model's usage for the description of high-temperature superconductivity \cite{yang89}.
Shortly afterward, an exact SO(4) symmetry was discovered, which allows a degeneracy of superconductivity (SC) and a charge-density wave (CDW) governed by an attractive on-site interaction \cite{yang90}.
This coexistence can be argued by the existence of a particle-hole transformation on bipartite lattices, 
that maps the attractive Hubbard model exactly onto the repulsive one, exhibiting antiferromagnetism (AFM) \cite{Hirsch85}.
The previously mentioned SC and CDW phases map to different spin operators \cite{zitko15,lieb89}.

Numerous studies have investigated the phases and various quantities of the Hubbard model in equilibrium systems, including various additional interactions with and without doping.
\cite{Micnas88,Micnas88b,Micnas89,Dzierzawa92,Kostyrko92,Eriksson95,Staudt00,Onari04,Toschi05,Brackett16,Paki19,romer20,Sushchyev22}.
Recent studies on the dynamics of the superconducting gap parameter examined quenches, yielding oscillations \cite{Volkov73,Yuzbashyan05,Yuzbashyan06,Barankov06,Cui19}, 
and driven systems in the hope of inducing superconductivity \cite{Nicoletti14,Krull14,Moor14,Casandruc15,patel16,sentef17,Buenemann17}.

In this paper, we will restrict ourselves to the half-filled Hubbard model, including an additional intersite interaction, on a square and a simple cubic lattice.
Specifically, we investigate competing phases, i.e., the behavior of various collective modes around phase transitions.
To this end, the 2D case already provides a wide range of possible phases, including CDW, AFM, $s$- and $d_{x^2 - y^2}$-wave superconductivity, and phase-separated states \cite{Micnas88b,Tsuchiura95,Su01,Su04,ha11,Huang13,Jiang22,Linner23}.
We expand these well-known phase diagrams to the simple cubic lattice where we find a similar structure for CDW, AFM, and $s$-wave SC.

@@

Moreover, we demonstrate an explicit way to obtain various Green's functions from the aforementioned methodology.
By extension, we also obtain the spectral functions of the investigated systems and discuss the excitations found therein.
The prominent examples in the SC phase are the well-known phase (Anderson-Bogoliubov) and amplitude (Higgs) modes.
The former occurs in neutral superfluids at $0$ energy and has been found in the past \cite{Maiti2015,Sun2020,Fan22}.
We will obtain this mode as well based on a microscopic description that excludes long-range electromagnetic interactions.
The inclusion of this kind of interaction shifts this mode towards the plasma frequency \cite{Kulik1981}. TODO: Schön

The amplitude mode ... \cite{Varma02,Cea14,Measson14,Tsuji15,Krull16,Schwarz20}
The latter is found at the lower edge of the two-particle continuum.

@@

In our study, we employ the extended Hubbard model as it encompasses the relevant phases and thereby provides direct access to the rich excitation spectra therein.
Its Hamiltonian is given by
\begin{equation}
    \label{eqn:full_hamiltonian}
    \begin{aligned}
        H = &-t \sum_{\langle i, j \rangle, \sigma} \left( c_{i\sigma}^\dagger c_{j\sigma} + \text{h.c.} \right) 
        + \mu \sum_{i,\sigma} n_{i\sigma} \\
        &+ U \sum_{i} n_{i\uparrow} n_{i\downarrow} 
        + \frac{V}{2} \sum_{\langle i, j\rangle, \sigma} n_{i\sigma} n_{j\sigma}\,,
    \end{aligned}
\end{equation}
where $c_{i\sigma}^{(\dagger)}$ annihilates (creates) an electron with spin $\sigma$ on lattice site $i$ 
and $\langle i, j\rangle$ denotes the summation over the next neighbors.
The parameters are the hopping amplitude $t$, the onsite interaction $U$, the intersite interaction $V$, and the chemical potential $\mu$.
Applying a Fourier transform into $k$ space yields the single-particle dispersion 
\begin{equation}
    \epsilon_0 (\vk) = -2t \sum_{\alpha=1}^D \cos(k_\alpha)\,,\quad k_\alpha \in [-\pi, \pi)\,,
\end{equation}
with the system's dimension $D$ and the dimensionless momentum $\vk$.

@@

In this article, we will be using the symplectic product\textcolor{red}{, also referred to as the operator pseudo scalar product,}
\begin{equation}
\label{eqn:scalar_product}
    (A | B) \coloneqq  \langle [A^\dagger, B] \rangle\,,
\end{equation}
where the expectation values are taken with respect to the thermal equilibrium of the mean-field Hamiltonian \eqref{eqn:mf_hamiltonian}.
Note, that this is not a proper scalar product, as a norm defined using it can also be negative.
For example, assume that for some operator $A$
\begin{equation}
    ||A|| = (A | A) = \langle [A^\dagger, A] \rangle > 0\,,
\end{equation}
then
\begin{equation}
    ||A^\dagger|| = (A^\dagger | A^\dagger) = \langle [A, A^\dagger] \rangle = - ||A|| < 0\,.
\end{equation}
Nevertheless, this does not pose any issue in our calculations.
We will compute $\mM$ by commutating with the full Hubbard Hamiltonian \eqref{eqn:full_hamiltonian} enabling us to capture collective behavior rather than the one particle dynamics given by the mean-field approximation.
